---
title: "Untitled"
output: html_document
---

```{r libraries, warning=FALSE,message=FALSE}
library(tidyverse)
library(magrittr)
library(goalmodel)

set.seed(3459)
```

Lets imagine a summer league between 6 English football clubs where each team plays each other twice (once at home and once away)

```{r teams, warning=FALSE,message=FALSE}
teams <- c("Arsenal", #5th in the 1st tier
           "Blackburn_Rovers", #15th in 2nd tier
           "Coventry_City", #8th in 3rd tier
           "Dover_Athletic", #14th 5th tier 
           "Enfield_Town", #10th in 7th tier
           "Frimley_Green") #2nd in 9th tier

```

We've managed to arrange a league that has a nice stratification between teams, so we'd expect each to be comfortably better than the next best (which will make sanity checking our results easier). Lucky for us, the teams are also in alphabetical order of strength so in case you don't have any prior on a team, take the first letter of it's name (A-F).

Each week each team play one game, so we'll have a fixture list that looks like:

```{r create_fixtures, warning=FALSE,message=FALSE,include=FALSE}
#https://stackoverflow.com/questions/54099990/is-there-an-efficient-algorithm-to-create-this-type-of-schedule
create_fixtures <- function(teams) {
  team1 <- as.character(teams[1])
  other_teams <- as.character(teams[!teams %in% team1])
  length <- length(other_teams)
  
  for(week in seq((length(teams)-1)*2)) {
    
    if(week %% 2 == 0) {
      fixtures <- data.frame(home = c(team1, other_teams[1:2]),
                             away = other_teams[length:3],
                             gameweek = week)
    } else {
      fixtures <- data.frame(home = other_teams[length:3],
                             away = c(team1, other_teams[1:2]),
                             gameweek = week)
      
    }
    
    if(week == 1) {
      fixtures_df <- fixtures 
    } else {
      fixtures_df <- rbind(fixtures_df, fixtures)
    }
    
    other_teams <- c(other_teams[length], other_teams[1:length-1])
  }
  
  return(fixtures_df)
}

fixtures <- create_fixtures(teams) %>%
  mutate_if(is.factor, as.character)
```
```{r create_results, warning=FALSE,message=FALSE,include=FALSE}
model <- list()
model$parameters <- list(attack = seq(1, -1 + 2/length(teams), by = -2/(length(teams)-1)) %>%
                           append(-sum(.)) %>%
                           `names<-`(teams), 
                         defense = seq(1, -1 + 2/length(teams), by = -2/(length(teams)-1)) %>%
                           append(-sum(.)) %>%
                           `names<-`(teams), 
                         intercept = 0, 
                         hfa = 0.3)

model$all_teams <- teams

model$model <- "poisson"
model$maxgoal <- 8

results <- predict_expg(model, fixtures$home, fixtures$away, return_df = TRUE) %>%
  mutate(noise1 = rnorm(nrow(.), 0, 0.6),
         noise2 = rnorm(nrow(.), 0, 0.6)) %>%
  mutate(hgoal = round(expg1 + noise1,0 ),
         agoal = round(expg2 + noise2,0),
         home = as.factor(team1),
         away = as.factor(team2)) %>%
  merge(., fixtures, by = c("home", "away")) %>%
  mutate_at(vars(hgoal:agoal), funs(replace(., .<0, 0))) %>%
  select(home, away, hgoal, agoal, gameweek) %>%
  arrange(gameweek, home) %>%
  filter(gameweek <= 8)

results
```

```{r show_fixtures, warning=FALSE,message=FALSE}
head(fixtures, 8)
```

Obviously for this we're going to have to make up our data. For the code used to generate it, see the bottom of the post.

Let's say that we've had 8 weeks of games played so far, and the results have been as follows

```{r print_results, warning=FALSE,message=FALSE}
head(results,8)
```

A better way to show this is to generte a matrix of home (y axis) vs. away (x axis) which gives us:

```{r plot_results, warning=FALSE,message=FALSE}
p1 <- results %>%
  filter(!is.na(hgoal)) %>%
  ggplot(., aes(x = away, y = home, fill = hgoal-agoal)) +
  geom_tile() +
  geom_label(aes(label = paste(hgoal, agoal, sep = "-")), fill = "white") +
  scale_fill_gradient2(low = "darkred", high = "green", midpoint = 0, guide = FALSE) +
  scale_x_discrete(limits = levels(results$home), position = "top") +
  scale_y_discrete(limits = rev(levels(results$away))) +
  theme_minimal()

p1
```

As the colour gradient (from bottom right to top left) shows, the teams we'd expect to do better are. Given the stochastic nature of football though, there are some surprises. E.g. Blackburn only managing to draw at home to Coventry.

A good sense of teams relative abilities can be seen in the league table of results so far (assuming 3 points for a win, and 1 for a draw):

```{r}
melt_results <- function(results_df) {
  results_df %>%
    select(home, away, hgoal, agoal) %>%
    gather(location, team,  -hgoal, -agoal) %>%
    #calculate goals for/against the team
    mutate(g_for = case_when(
      location == "home" ~ hgoal,
      location == "away" ~ agoal
    )) %>%
    mutate(g_ag = case_when(
      location == "home" ~ agoal,
      location == "away" ~ hgoal
    )) 
}

results_to_table <- function(results_df) {
  results_df %>%
    melt_results(.) %>%
  #3 points for a win, 1 for a draw
  mutate(points = case_when(
    g_for > g_ag ~ 3,
    g_ag > g_for ~ 0,
    g_for == g_ag ~ 1
  )) %>%
  #calculate goal difference for each match
  mutate(gd = g_for - g_ag) %>%
  group_by(team) %>%
  #get the final statistics per team
  summarise(games_played = n(),
            gf = sum(g_for),
            ga = sum(g_ag),
            gd = sum(gd),
            points = sum(points)) %>%
  arrange(-points, -gd, -gf)
}

league_table <- results  %>%
  filter(!is.na(hgoal)) %>%
  select(-gameweek) %>%
  results_to_table(.) %>%
  print()

```

Where teams positions are nicely rank ordered.

However, with two games to play each, there's an interesting

However, this still leave 

```{r, include = FALSE}
model <- goalmodel(results$home, results$away, results$hgoal, results$agoal)

params <- model %>%
  #select team parameters
  .[grepl("defense|attack", names(.))] %>%
  do.call(cbind, .) %>%
  as.data.frame() %>%
  rownames_to_column("team")

#munge the other parameters into text
#the home field advantage and the intercept
other_params <- model %>%
  .[!grepl("defense|attack", names(.))] %>%
  as.data.frame() %>%
  mutate_if(is.numeric, round, 3) %>%
  #glue into one object
  glue::glue_data("intercept: {.$intercept},\n hfa: {.$hfa}")

#plot team paramters over attack/defense axis
p2 <- ggplot(params, aes(x = attack, y = defense, label = team)) +
  #add a best fit linear regression line
  stat_smooth(method = "lm", se = FALSE) +
  geom_point() +
  ggrepel::geom_text_repel(nudge_y = 0.1) +
  #annotate the other parameters in
  annotate("text", y = min(params$defense), x = max(params$attack)*0.6, label = other_params, hjust = 0) +
  theme_minimal()

p2

```
